Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-28 demystify.tex
updates

    % Title
    \node[align=center] at ($(current page.center)+(0,-5)$) 
    {
    {\fontsize{24}{24} \selectfont {{Demystifying the Computer Science}}} \\[0.15in]
    {\fontsize{24}{24} \selectfont {{PhD Admission in the US}}} \\[0.2in]    
    {\fontsize{18}{18} \selectfont {{A Handbook for International and Domestic Students}}} \\[0.5in]

@@

\chapter*{Preface}
Having been involved in PhD admission committees for many years, I've realized that many \textbf{international} (and domestic) students, especially those in smaller countries or less well-known universities, lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted to a CS PhD program in the US is difficult compared to other countries.

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this handbook is primarily intended for students interested in CS, it might be relevant to students from various disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) US universities are classified as R1.} institutions in the US (and universities in other countries).

@@

    \item Miscs and FAQS
        \begin{itemize}
          \item Why you were rejected (\autoref{sec:why-rejected})
          \item Increasing admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item Being in an underrepresented group can help can help (\autoref{sec:urm}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\autoref{sec:non-stem}).
          \item You do not need an MS degree (\autoref{sec:msrequirement}) to apply for PhD in CS, and it takes about 5--7 years for PhD in CS (\autoref{sec:time}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\autoref{sec:non-us-differences}).
          \item How do you call a professor? (\autoref{sec:address})
          \item Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it (\autoref{sec:ra-cost})
          \item Ask your prof. if they can buy computer equipments and such for your research (\autoref{sec:buying-equipment}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
        \end{itemize}
\end{enumerate}

@@

Students, especially international ones, often wonder if they should apply to a CS PhD program in the US. Specifically, they are worried about the \emph{difficulty} of getting admitted and the \emph{cost} of studying in the US. This section addresses these concerns.

@@

\section{Accepting, Postponing, and Decline Offers}\label{sec:accept-postpone-decline}

Students often ask about what to do after they get an offer from a professor, e.g., is it OK to postpone or accept/reject offers?, and most importantly, how to do so without offending anyone. Many also wonder about accepting an offer and then later rejecting it.

In short, if you are not serious about an offer for any reason (e.g., still waiting for other offers), \emph{do not accept it}.
If you are not sure, ask for more time to decide.  But do not accept an offer and then later reject it.  This can be considered \emph{rude and unprofessional}, and will likely burn bridges with that prof.

@@

% If you have better offers, you might mention them to the professor and ask if they could match them. It is unlikely they can do so as RA might be fixed by the department and not negotiable. They might recommend you for departmental fellowships or some additional perks.

@@

This is very similar to a professional job offer.  You should be serious about it and communicate promptly. Importantly, do not accept an offer and then later reject it.  If you do this with a company, you might face legal consequences.  If you do this with a professor, they will likely not do anything about it (i.e., not worth their time) but they will not want to work with you in the future.

% 1. I'm fine if I extend an offer and you request a delay, even if it's because you're exploring other programs. No issue here, it's similar to job offerings. Of course, I may not be able to wait indefinitely and may extend the offer to another student, but I'll let you know that.
% 2. I'm also okay if you find a program or professor that's a better fit. Similar to the first point, no hard feelings at all. However, I'd appreciate a prompt notification of your decision so I can extend an offer to another student.
% 3. I'm NOT okay if you accept my offer but then, upon finding something better, tell me of your decision to withdraw Just like in job offers, once you are committed to an offer, there should be a compelling reason to withdraw (and claiming "I found somewhere better" is the weakest excuse).
% While I won't take any actions against you (a company might), I naturally will remember this and may be cautious in future interactions. I'll also reconsider making similar offers to students in similar situations.

>>>>

    % Title
    \node[align=center] at ($(current page.center)+(0,-5)$) 
    {
    {\fontsize{24}{24} \selectfont {{Demystifying the Computer Science}}} \\[0.15in]
    {\fontsize{24}{24} \selectfont {{PhD Admission in the US}}} \\[0.2in]    
    {\fontsize{18}{18} \selectfont {{A Handbook for International Students}}} \\[0.5in]

@@

\chapter*{Preface}
Having been involved in PhD admission committees for many years, I've realized that many \textbf{international} students, especially those in smaller countries or less well-known universities, lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted to a CS PhD program in the US is difficult compared to other countries.

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
Originally, this document was intended for international students, but I have expanded it to include information that might be useful for \emph{domestic students} as well.
While this handbook is primarily intended for students interested in CS, it might be relevant to students from various disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) US universities are classified as R1.} institutions in the US (and universities in other countries).

@@

    \item Miscs and FAQS
        \begin{itemize}
          \item Why you were rejected (\autoref{sec:why-rejected})
          \item Increasing admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item Being in an underrepresented group can help can help (\autoref{sec:urm}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\autoref{sec:non-stem}).
          \item You do not need an MS degree (\autoref{sec:msrequirement}) to apply for PhD in CS, and it takes about 5--7 years for PhD in CS (\autoref{sec:time}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\autoref{sec:non-us-differences}).
          \item How do you call a professor? (\autoref{sec:address})
          \item Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it (\autoref{sec:ra-cost})
          \item Ask your prof. if they can buy computer equipments and such for your research (\autoref{sec:buying-equipment}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
        \end{itemize}

    \item Appendices
    \begin{itemize}
      \item Specific advice for domestic students (\autoref{sec:domestic-students}).
      \item How to find research opportunities (\autoref{sec:research-opportunities}).
      \item Cultural differences between US and other countries (\autoref{sec:cultural}).
    \end{itemize}

\end{enumerate}

@@

Students, especially those from smaller countries or less well-known universities, often wonder if they should apply to a CS PhD program in the US. Specifically, they are worried about the \emph{difficulty} of getting admitted and the \emph{cost} of studying in the US. This section addresses these concerns.

@@

\section{Checking Status, Accepting, Postponing, and Decline Offers}\label{sec:accept-postpone-decline}

Students often ask about what to do after they get an interview or an offer from a professor, e.g., if they can followup to find out about their status, or is it OK to postpone or accept/reject offers?, and most importantly, how to do so without offending anyone.

\paragraph{Checking your application status} If you have interviewed and not heard back from a professor after a few weeks or especially around the time when universities send out their admission decisions (around late Feb-- mid Mar), you can email to check.  You can follow up the interview invitation and say: \emph{``Thanks for chatting with me. I am very excited about the opportunity to work with you.  Could you please let me know if you have made a decision or if you need more information from me?''}.  If you have new updates, e.g., new publications or new fellowship awards, or even new offers from other professors or schools, you can also mention that.

Profs. are often very busy, especially during admission time when they have to review many applications, interview many students, and make decisions.  They might not have time to respond to every email.  If you do not hear back after a week, you can send another email to check again.  If you still do not hear back, you can assume that you are not selected.

@@

@@

\paragraph{Accepting an offer and later later rejecting it}
I have seen \emph{many} international students asking about this, e.g., accepting an offer from a prof. and then later rejecting it when they get admitted to a better/dream school, which have them on its wait list. There are numerous advice, e.g., on Internet forums, that this is OK to do so (most of these advice are from other international students).
The reasons typically are \emph{``you haven't established much relationship with the prof. yet, so it is OK to reject it; no hard feeling, you probably won't see them again anyway''} or that \emph{``it's your life and you should do what's best for you''}.

From my experience, this is \emph{not} a good idea and can be considered \emph{rude and unprofessional}. But ultimately, it is a trade-off that you need to decide,  e.g., between your own benefit and the cost to others. If you reject an offer, the prof. will likely stop looking for other students and might even reject other students; so by rejecting the offer later, you causing a great deal of inconvenience to the prof and also taking away the opportunity from other students. This will likely would burn bridges with that prof and they will remember what you did to them.  Moreover, given that research subfields are typically small, you might run into that professor again in the future, and they will remember what you did to them.
